---
title: "DATA2004 Refresher"
subtitle: "Tidyverse + Quarto Review"
author: "Danielle Pruitt"
format: html
editor: visual
---

## Before you begin

The goal of this refresher is to practice the R skills you learned in DATA1004 and get more comfortable working in a Quarto document.

This is **not new material**. You should look through your notes and materials from DATA1004 and the R4DS book for this course if you get stuck. And **please** feel free to email me if you want any help at all.

As you work:

1.  Write your R code inside code chunks.
2.  Write your answers and interpretations as normal text **outside** code chunks.
3.  Render your document periodically rather than waiting until the end.
4.  Your final document should render successfully to HTML.

Please refer to this [Quarto guide](https://quarto.org/docs/get-started/hello/rstudio.html) for a helpful review of Quarto.

Please refer to Dr. LeVan's [Book](https://levanjm.github.io/Book-2/) for additional review of R.

You will **not** be graded for this. I realized today it's been over a year since some of us have worked in R at all, so this is to give you a chance to see what you remember, give you a chance to review, and give you the opportunity to become reacquainted with Quarto. :)

There are gaps before headings for you to put your code chunks and written responses.

## Setup

First, let's load the packages we need.

You'll need to run `install.packages(palmerpenguins)` **once** in your console if you haven't done so before.

```{r}
library(tidyverse)
library(palmerpenguins)
```

We will use the `penguins` dataset from the `palmerpenguins` package.

```{r}
penguins <- penguins
glimpse(penguins)
```

### Getting oriented

Two questions:

1.  What are the dimensions of the `penguins` dataset? The penguins dataset is a 8X344 dataset.
2.  What does one row of `penguins` seem to represent? One row of penquins represents a variable about each penguin that we're interested in.

## Getting reacquainted with the basic verbs

### Select

Create an object called `penguins_small` that contains only:

1.  species

2.  island

3.  sex

4.  body_mass_g

5.  flipper_length_mm

    ```{r}
    penguins_small <- select(penguins, "species", "island", "sex", "body_mass_g", "flipper_length_mm")
    ```

### Filter

Using the original `penguins` dataframe, create a dataframe containing only **Adelie** penguins using `filter().`

```{r}
adelie_penguins <- filter(penguins, species == "Adelie")
```

Using the original `penguins` dataframe, find all penguins with a body mass greater than **5,000 grams** using `filter`.

```{r}
big_penguins <- filter(penguins, body_mass_g > 5000)
```

And then use `glimpse()` to look at your results for all of the above.

```{r}
glimpse(adelie_penguins)
glimpse(big_penguins)
```

### Creating new variables

Create a new dataframe with a variabled called `body_mass_kg` that converts each penguin's body mass in kilograms.

```{r}
penguins_mutated <- mutate(penguins, body_mass_kg = body_mass_g / 1000)
```

# 

## Summaries

Let's say I want you to compare three penguin species, but I don't tell you which three in my email.

First, find out the different values in `species` and then select any three (since I didn't think to email you again and left my office) and give me summaries of

1.  The number of observations;

    ```{r}
    unique(penguins$species)

    penguins_Gentoo <- filter(penguins, species == "Gentoo")
    penguins_Chinstrap <- filter(penguins, species == "Chinstrap")


    ```

    There are 152 observations of Adelie, 124 observations of Gentoo, and 68 observations of Chinstrap.

2.  mean body mass;

    ```{r}
    adelie_penguins <- filter(adelie_penguins, body_mass_g != "NA")
    mean(adelie_penguins$body_mass_g)

    penguins_Gentoo <- filter(penguins_Gentoo, body_mass_g != "NA")
    mean(penguins_Gentoo$body_mass_g)

    penguins_Chinstrap <- filter(penguins_Chinstrap, body_mass_g != "NA")
    mean(penguins_Chinstrap$body_mass_g)
    ```

    The mean body mass of Adelie penguins is 3700.662 g. The mean body mass of Gentoo penguins is 5076.016g. The mean body mass of Chinstrap penguins is 3733.088g.

3.  median body mass

    ```{r}
    median(adelie_penguins$body_mass_g)
    median(penguins_Gentoo$body_mass_g)
    median(penguins_Chinstrap$body_mass_g)
    ```

    The median body mass of Adelie penguins is 3700g. The median body mass of Gentoo penguins is 5000g. The median body mass of Chinstrap penguins is 3700g.

4.  mean flipper length

    ```{r}
    mean(adelie_penguins$flipper_length_mm)
    mean(penguins_Gentoo$flipper_length_mm)
    mean(penguins_Chinstrap$flipper_length_mm)
    ```

The mean flipper length for Adelie penguins is 189.95mm. The mean flipper length for Gentoo penguins is 217.19mm. The mean flipper length for Chinstrap penguins is 195.82mm.

*note:* some could have missing values. how should you deal with that?

## More independent review

From here on, I will tell you the result I want and then it is up to you to figure it out.

### A)

Among **female** penguins, which species has the greatest average body mass?

```{r}
female_adelie_penguins <- filter(adelie_penguins, sex == "female")
mean(female_adelie_penguins$body_mass_g)

female_Gentoo_penguins <- filter(penguins_Gentoo, sex == "female")
mean(female_Gentoo_penguins$body_mass_g)

female_Chinstrap_penguins <- filter(penguins_Chinstrap, sex == "female")
mean(female_Chinstrap_penguins$body_mass_g)
```

Female Gentoo penguins have the largest average body mass of the 3 species.

### B)

For each species and island combination, determine

1.  how many observations are present;
2.  mean flipper length

```{r}
install.packages("gt")
library(gt)

summary_table <- penguins %>%                             # Step 1
  group_by(sex, island) %>%                        # Step 2
  summarise(                                        # Step 3
    avg_flipper_length = mean(flipper_length_mm, na.rm = TRUE),
    total_penguins = sum(count, na.rm = TRUE)
  ) %>%
  arrange(desc(total_penguins)) %>%                  # Step 4
  gt() %>%                                          # Step 5
  tab_header(                                       # Step 6
    title = "Penguin Flipper Length Summary by Sex and Island",
    subtitle = "Average Flipper Length and Total Observations"
  ) %>%
  fmt_number(                                       # Step 7
    columns = c(avg_flipper_length), 
    decimals = 2
  ) %>%
  fmt_number(
    columns = c(total_penguins), 
    decimals = 0,
    use_seps = TRUE            # This adds separators (commas) for thousands
  ) %>%
  cols_label(                                       # Step 8 
    sex = "Penguin Sex",
    island = "Island",
    avg_flipper_length = "Average Flipper Length (mm)",
    total_penguins = "Total Penguins"
  ) %>%
  cols_align(                                       # Step 9
    align = "center",
    columns = c(sex, island, avg_flipper_length, total_penguins) # Specify columns to center
  )
```

It would be very helpful to review making tables.

Then arrange the resulting table from **greatest to smallest** mean flipper length

### C)

Which individual penguins have the five greatest recorded body mass? Your result should contain only:

1.  species
2.  island
3.  sex
4.  body mass

## Viz review

1.  Create a visualization that compares the **distribution of body mass across species.**

    1.  Give your plot a meaningful title and label your axes.

        ```{r}
        ggplot(penguins, aes(x = species, y = body_mass_g)) +
          geom_col() +
          labs(x = "Species", y = "Body Mass (g)",
               title = "Distribution of Body Mass by Penguin Species")
        ```

2.  Create a scatterplot showing the relationship between flipper length and body mass.

    1.  Create a new scatterplot for this relationship, but differentiate species by any aesthetic of your choosing.

    ```{r}
    ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
      geom_point(color = species) +
      labs(x = "Flipper Length (mm)", y = "Body Mass (g)",
           title = "Fipper Length vs. Body Mass by Species")
    ```

    1.  Create a new scatterplot for this relationship, but only include **three** species and then facet to examine the relationship across three species separately.

```{r}
three_species <- filter(penguins, species == "Adelie",  "Gentoo", "Chinstrap")

ggplot(three_species, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point() +
  labs(x = "Flipper Length (mm)", y = "Body Mass (g)",
       title = "Fipper Length vs. Body Mass by Species") +
  facet_wrap(~species)
```

## Put it together

Hopefully now you've done enough of the review to get a completely open-ended question. For your last question, here's what id like to see

Do male and female penguins appear to differ in body mass, and does that difference look similar across species?

It would be very helpful to review how to single out individual variables. I'm not exactly sure where to start with this question.
